Curation, filtering, and dedup make real corpora missing-not-at-random: they
censor whole regions of input space, not just thin them. Reweighting fixes
(Heckman, IPW, counterfactual-risk-minimization) can only correct the observed
points --- they cannot place mass where there are zero observed samples. We ask
whether, once the selection mechanism is estimated, one can \emph{manufacture}
the censored region rather than only reweight around it. We estimate
$s(x)=P(\text{observed}\mid x)$ by a calibrated density-ratio deep ensemble
(recovering a known 2D selector globally and on the complement, Spearman
$\ge\EZeroRhoCompMin$, ECE $\le\EZeroEceMax$), and steer a Doob-$h$-transform
diffusion bridge whose guidance signal is the \emph{estimated selection
propensity} $1-\shat$ --- to our knowledge the only method that uses a selection
function to \emph{locate and synthesize whole new units} in a never-observed
feature region, rather than reweighting or imputing labels for observed units.
The bridge works \emph{as a generator}: it steers synthesis into the censored
complement roughly twice as often as an identically-strong misdirected control
(complement-hit $\EOneCompHitMethod$ vs $\EOneCompHitDecoy$) while keeping
off-manifold rejection under $\EOneRejectMethod$. Yet our pre-registered
discriminator --- the method-minus-\emph{misdirected}-guidance gap on the
censored slice as a function of selection entropy $\IOX$ --- is a
\textbf{negative}: the gap does not rise with $\IOX$ (Spearman $\KOneSpearman$,
$p{=}\KOneSpearmanP$), the method does not exceed the matched decoy at strong
censoring ($p{=}\KTwoMethodVsDecoyP$), and it underperforms $1/\shat$ reweighting
by $\MethodVsBThreeGap$ points. A central-squeeze analysis explains why, and this
is the paper's contribution: the region where $\shat$ is trustworthy --- the
short-extrapolation collar the identifiability gates confine generation to ---
is exactly where reweighting already succeeds, while the recoverable headroom
lies in the deep complement where $\shat$ is off-support extrapolation and
steering is (correctly) forbidden. Selection-guided generation thus buys, at
most, a bounded collar of advantage whose width is set by $\shat$ extrapolation
reliability, and on our testbeds that collar is too narrow to beat reweighting.
We report the negative honestly with its characterization, provide the
left-end-zero theorem and a foreign-structure positive control, and do not claim
the synthetic result transfers to real corpora.
