This section is the honesty of the paper; each item is a claim we explicitly do
\emph{not} make.

\paragraph{Identifiability.} $s(x)=P(\text{observed}\mid x)$ is not
pointwise-identified in the censored region: there $\Dobs$ has zero points, so
$\shat$ is classifier off-support extrapolation, not a measurement. The
recoverable deliverable is the observed-support \emph{collar} (density- and
uncertainty-gated) --- censored \emph{combinations} of present structure, not
arbitrary deep-complement structure.

\paragraph{The central squeeze.} Where generation provably beats $1/\shat$
reweighting (B3) is \emph{beyond} the collar, exactly where $\shat$ is least
reliable; where $\shat$ is reliable, B3 nearly matches. The contribution is a
\emph{bounded} collar of advantage whose width is set by $\shat$ extrapolation
reliability --- we own this as the central finding (Fig.~\ref{fig:squeeze}), not
a blowout.

\paragraph{The theorem is a bound plus an empirical hypothesis.} Only the left
end (gain $\to 0$ as $\IOX\to 0$) is proven; the rise is empirical and, on its
own, nearly trivial, so we do not state ``gain tracks selection entropy'' as a
theorem. The discriminating evidence is the method-minus-misdirected-guidance
gap.

\paragraph{The Doob bridge is approximate.} Reconstruction guidance and twisted
SMC both ride the biased $r(\hat x_0)$ proxy; we describe the procedure as
``guided generation toward an estimated-low-selection collar,'' not exact Doob
conditioning.

\paragraph{Scope fence.} The method recovers only structure on the manifold
spanned by the observed data (interpolative, censored combinations); it does not
generate foreign structure. Our K4 probe is a positive control that the method
\emph{must} fail on a genuinely non-recombinable censored mode, and it does
(\S\ref{sec:scope}).

\paragraph{Feasibility/realism gap.} The cheap decisive experiment is synthetic,
and its advantage partly rests on a full-support reference cover --- a controlled
affordance not available on real corpora. The real-corpus rung
(\S\ref{sec:real}) is assumption-gated by a cover-blind check and is
limits-characterization only; a synthetic go/no-go must never be read as a
real-corpus go.
