\paragraph{Real-corpus pilot (limits only).} As a limits check we curate a real
tabular corpus --- California housing, two standardized features, target
above-median value --- with a \emph{real} covariate driving the MNAR selection
($s_\beta$ on standardized median income, censoring high-income tracts), taking
the raw unfiltered pool as the reference cover. A cover-blind check confirms the
reference is not co-censored on the target region (the selector retains signal
there, slice $\shat$ spread $\EFiveCoverStd$; K5 not triggered). At strong
censoring the method attains $\EFiveSliceMethod$ slice accuracy versus
$\EFiveSliceBZero$ for the obs-only floor and $\EFiveSliceDecoy$ for the
misdirected decoy --- and here it \emph{does} beat $1/\shat$ reweighting
($\EFiveSliceBThree$), which becomes high-variance when a few heavily-censored
tail points carry enormous weights. The lesson is not that generation wins: the
method-minus-\emph{decoy} margin --- the selection-targeting signal --- remains
within noise, exactly as in 2D. It is that generation's advantage over
\emph{reweighting} is dataset-dependent (it appears precisely where reweighting
is ill-behaved), whereas its advantage over matched misdirected guidance is the
invariant, and invariantly small. We report this as characterization, not a
real-corpus go: three seeds, a synthetic-style selector on real features, and a
full-support reference cover a genuine corpus would not hand us.

\paragraph{The perceptual rung.} A faithful perceptual testbed (pixel-space
subpopulation censoring with an EDM/twisted-SMC bridge) is the natural next rung.
A cheap latent-space proxy we ran (PCA of MNIST with the same pipeline) was
uninformative on both ends --- the digit subpopulation the thickness selector
censors remains easily classifiable, so no downstream headroom exists to
contest, and a small latent diffusion is flagged off-manifold by the independent
validator --- so we do not report it as evidence and leave the faithful
pixel-space version, which the 2D and real-tabular findings already frame, to
future work.
