\documentclass[10pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue!60!black,citecolor=blue!60!black,urlcolor=blue!60!black]{hyperref}
\usepackage{caption}
\captionsetup{font=small,labelfont=bf}

\input{numbers}

\newcommand{\loss}{\mathcal{L}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}

\title{Rice's Formula for Event-Driven Networks:\\
Predicting and Training Temporal Sparsity via\\ Differentiable Level-Crossing Rates}

\author{Gunner Levi Howe\\
Independent Researcher\\
\texttt{gunnerlevihowe@gmail.com}}

\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
% FINAL ABSTRACT WRITTEN AFTER RESULTS -- honest, states what did not work.
\end{abstract}

\section{Introduction}

Delta networks---delta-RNNs, sigma--delta networks, event-driven
transformers---spend compute only when an activation \emph{changes} by more
than a threshold~$\theta$: each supra-threshold change transmits a delta,
triggers the corresponding weight-column MACs and memory traffic, and is
therefore a quantum of energy \citep{neil2017delta,oconnor2017sigmadelta,
yousefzadeh2022delta,gao2020edgedrnn,chen2024deltakws}. In this literature
the thresholds are tuned by post-hoc grid search and the event counts are
always \emph{measured after the fact}; no published delta-network method
predicts event rates from activation statistics or trains against an
analytic rate objective.

A delta event is, definitionally, a level crossing of the activation time
series. Level-crossing statistics have an analytic theory that is a century
old: Rice's formula \citep{rice1944,kac1943} gives the mean crossing rate of
a stationary Gaussian process at level $u$ from two spectral moments, and
its Monte-Carlo/co-area estimator---validated as a differentiable training
objective in our preceding program
\citep{howe2026levelcrossing,howe2026budget,howe2026minkowski}---needs no
Gaussianity at all. This paper connects the two: it gives the delta-network
literature the analytic, differentiable event-rate model it has been
missing, and it gives the level-crossing program the domain where excess
crossings are the cost \emph{by construction}.

% narrative arc paragraph (mechanism validated, pathology absent, native
% domain found) + contributions list -- FILLED AFTER RESULTS.

\section{Related work}
\label{sec:related}

\paragraph{Delta and sigma--delta networks.}
\citet{neil2017delta} threshold changes of RNN inputs and hidden states
against an anchor memory of the last transmitted value and accumulate
matrix--vector products incrementally; the mechanism transfers to
convolutional and embedded video settings \citep{yousefzadeh2022delta}, to
silicon \citep{gao2020edgedrnn,chen2024deltakws}, and underlies sigma--delta
quantized networks \citep{oconnor2017sigmadelta}. Throughout, thresholds are
set by accuracy-vs-speedup sweeps. EGRU \citep{subramoney2023egru} trains
activity-sparse GRUs with surrogate gradients through a hard threshold;
that is spike-style training, not an analytic rate model.

\paragraph{Spiking transformers and spatial activation sparsity.}
Spikformer \citep{zhou2023spikformer} and SpikeGPT \citep{zhu2023spikegpt}
use spike \emph{coding}---a different mechanism from delta thresholding; we
cite and do not claim their ground. The activation-sparsity line
\citep{li2023lazy} concerns \emph{spatial} sparsity (which units are zero);
ours is \emph{temporal} sparsity (which units change).

\paragraph{Level-crossing theory in ML and signal processing.}
Rice's formula appears in machine learning in loss-landscape critical-point
counting \citep{bray2007,choromanska2015} and in our own program
\citep{howe2026levelcrossing,howe2026budget,howe2026minkowski}---never as an
activation-event or energy model. Zero-crossing statistics are classical
discrimination features \citep{kedem1986,kedem1994book}; send-on-delta
sampling theory \citep{miskowicz2006} bounds event rates by total variation.
The novelty here, as in the companion papers, is the \emph{instantiation}:
crossing rates as a differentiable model of, and budget on, delta-network
energy.

\section{Theory: delta events as level crossings}
\label{sec:theory}

\paragraph{Setting.}
Let $a_c(t)$, $t=1,\dots,T$ be the sampled activation trace of channel $c$
feeding a delta-encoded matrix--vector product. The send-on-delta encoder
\citep{neil2017delta,miskowicz2006} keeps an anchor $\hat a$ (last
transmitted value), fires when $|a_c(t)-\hat a|>\theta$, and updates the
anchor to $a_c(t)$. Every event costs the fan-out of the weight column in
MACs plus the associated memory traffic (Section~\ref{sec:energy}).

\paragraph{Rice's formula and its discrete-time form.}
For a stationary process with spectral moments $\lambda_0=\operatorname{Var}
a$ and $\lambda_2=\operatorname{Var}\dot a$, the mean rate of up- and
down-crossings of level $u$ is
\begin{equation}
\mu(u) \;=\; \frac{1}{\pi}\sqrt{\frac{\lambda_2}{\lambda_0}}
\exp\!\Big({-\frac{(u-m)^2}{2\lambda_0}}\Big),
\label{eq:rice}
\end{equation}
under Gaussianity \citep{rice1944}. Sampled traces are sequences, not
continuous paths: for a stationary Gaussian \emph{sequence} with lag-1
autocorrelation $\rho_1$ the exact per-step crossing probability is
\begin{equation}
P\big[(a_t-u)(a_{t+1}-u)<0\big] \;=\; 4\,
\mathrm{T}\!\Big(\tfrac{u-m}{\sqrt{\lambda_0}},
\sqrt{\tfrac{1-\rho_1}{1+\rho_1}}\Big),
\label{eq:discrete}
\end{equation}
with $\mathrm{T}$ Owen's T function \citep{owen1956}; at $u=m$ this is the
classical $\arccos(\rho_1)/\pi$ \citep{kedem1994book}, and $\rho_1\to 0$
recovers the iid baseline $2\Phi(\tilde u)(1-\Phi(\tilde u))$.
Eq.~\eqref{eq:rice} with finite-difference moments underpredicts
Eq.~\eqref{eq:discrete} as $\rho_1\to0$ (by up to $\sim$10\%, quantified in
Section~\ref{sec:e0}); we use Eq.~\eqref{eq:rice} as the headline Gaussian
predictor, as specified, and Eq.~\eqref{eq:discrete} in the appendix.

\paragraph{Differentiable estimator.}
The program's smoothed Monte-Carlo estimator
\citep{howe2026levelcrossing} evaluates the co-area integrand
$\delta_\epsilon(a-u)\,|\dot a|$ at sample points. On a sampled trace, the
piecewise-linear segment $a_t\!\to\!a_{t+1}$ admits the \emph{exact}
smoothed count
\begin{equation}
\hat c(u) \;=\; \frac{1}{(T-1)\,\Delta t}\sum_{t=1}^{T-1}
\Big|\,\Phi\big(\tfrac{a_{t+1}-u}{\epsilon}\big)-
\Phi\big(\tfrac{a_{t}-u}{\epsilon}\big)\Big|,
\label{eq:segment}
\end{equation}
which converges to the hard sign-change count as $\epsilon\to0$, remains
differentiable in the trace values---and hence in the network
parameters---and is robust to steps that jump many $\epsilon$ in one
$\Delta t$ (where the midpoint co-area rule undercounts;
Section~\ref{sec:e0}).

\paragraph{From crossings to delta events.}
Summing crossing rates over the implicit ladder of transmission levels
spaced $\theta$ apart bounds the send-on-delta event rate by total
variation \citep{miskowicz2006}:
\begin{equation}
R(\theta)\;\le\;\min\!\big(\E|\Delta a|/\theta,\;1/\Delta t\big),
\qquad \E|\dot a| = \sqrt{2\lambda_2/\pi}\ \text{(Gaussian)}.
\label{eq:bound}
\end{equation}
The bound is loose by exactly the hysteresis of the anchor (re-crossing a
level without traveling $\theta$) and by multi-level jumps. We measure this
gap once on synthetic processes as a correction factor
$\gamma(\theta/\sigma_\delta)$, $\sigma_\delta$ the one-step increment
scale, and find it strikingly process-independent for
$\theta \lesssim \sigma_\delta$ (Section~\ref{sec:e0}).

\paragraph{One-sided crossing budget.}
The training objective transplants the howe2026budget mechanism to its
native domain: per layer $\ell$ with hidden traces $h_\ell$,
\begin{equation}
\loss_{\mathrm{budget}}^{(\ell)} \;=\;
\operatorname{mean}_j\;
\frac{\mathrm{relu}\big(\hat c_{\ell}(u_j) - b_j\big)^2}
{(b_j + \overline{b})^2},
\qquad b_j = \rho\, c^{\mathrm{cal}}_{\ell}(u_j),
\label{eq:budget}
\end{equation}
with levels $u_j$ on a quantile ladder of the layer's calibration
activation distribution (dense-ladder rule of \citet{howe2026minkowski}),
budgets a $\rho$-scaled copy of the calibration profile, and $\hat c$ the
segment estimator \eqref{eq:segment}. Within budget the loss and its
gradient are exactly zero (one-sidedness), so compliant dynamics are
untouched. The sampling-scale failure mode of \citet{howe2026minkowski}
does not arise here: traces are densely sampled 1-D sequences.

% Sections E0-E4, energy accounting, limitations, reproducibility --
% FILLED AFTER RESULTS.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
