% Single source of truth for numeric claims in the paper.
% Filled from results/ (see scripts/analyze.py output and results/*.csv).
% Char-LM numbers final (2026-07-02); synthetic prose updated after the sweep.

\newcommand{\resSynthSeeds}{three}
\newcommand{\resCharlmSeeds}{two}
\newcommand{\resCharlmParams}{11M}
\newcommand{\resSemrfOverhead}{2\%}

% ---------------------------------------------------------------- abstract --
\newcommand{\resAbstractFindings}{%
On enwik8, SemRF attains the best test bits-per-character of all eight schemes
at the training context (1.4246 vs.\ 1.4276 for RoPE, 1.4348 for ALiBi, and
1.4409 for CABLE) and, unlike rotary and absolute schemes---whose loss degrades
severely beyond the training length---its bpc \emph{continues to improve} with
longer evaluation contexts, reaching 1.340 at $8\times$ the training length,
ahead of CABLE (1.352) and ALiBi (1.356) with a margin that widens as context
grows. On the diagnostics, SemRF is the \emph{only} scheme that solves all
three tasks at the training length: RoPE and the absolute schemes fail
associative recall, ALiBi forms its retrieval circuit in only two of three
seeds, and CABLE---while excellent on recall---fails selective copying
outright. Ablations attribute SemRF's recency and recall behaviour causally to
the frame-conditioned temporal decay, while the comparison with CABLE
delineates an honest trade-off between structured and unstructured
content-conditioned biases on tasks requiring scale-free retrieval across
arbitrary gaps.}

% ----------------------------------------------------------------- results --
\newcommand{\resResultsProse}{%
\paragraph{Character-level language modeling.}
Table~\ref{tab:charlm} reports test bpc at the training context length (512).
SemRF achieves the best bpc of all eight schemes ($1.4246 \pm 0.0006$ over
\resCharlmSeeds{} seeds), ahead of RoPE ($1.4276 \pm 0.0039$), ALiBi
($1.4348 \pm 0.0004$), and CABLE ($1.4409 \pm 0.0019$), at a parameter
overhead of only 1.7\% and identical backbones. The ordering is consistent
across seeds.

\paragraph{Length extrapolation on enwik8.}
Figure~\ref{fig:charlm-extrap} shows test bpc as the evaluation context grows
from 256 to 4096 with no retraining. The absolute schemes fail catastrophically
(sinusoidal reaches 35 bpc at 4096; learned-absolute 4.1), and RoPE more than
triples its loss (1.43$\,\to\,$4.62). The T5 bucketed bias holds up to
$2\times$ before degrading. The three decay-based biases improve
monotonically with context: ALiBi from 1.435 to 1.356, CABLE from 1.441 to
1.352, and SemRF from 1.425 to \textbf{1.340} (means over seeds)---SemRF is
best at every evaluated length. Notably, SemRF's advantage over ALiBi
\emph{widens} with context length ($-0.010$ bpc at 512 and 1024, $-0.013$ at
2048, $-0.016$ at 4096): both content-conditioned biases (CABLE and SemRF)
out-extrapolate ALiBi's fixed slopes, and SemRF's structured
frame-conditioning outperforms CABLE's unstructured per-token conditioning
throughout.

\paragraph{Diagnostics: temporal recency.}
On the time-sensitive task (Table~\ref{tab:synth}, Figure~\ref{fig:synth-extrap}
left), only the decay-based biases are reliable: SemRF, ALiBi, and CABLE all
reach perfect accuracy on the training distribution ($1.000 \pm 0.000$ over
three seeds) \emph{and remain perfect at $4\times$ longer streams}.
Learned-absolute embeddings also solve the training length ($1.000$) but
collapse under extrapolation ($0.66$ at $2\times$, $0.36$ at $4\times$)---the
canonical absolute-position saturation failure. RoPE is bistable across seeds
($0.60 \pm 0.28$ at train length; one of three seeds solves the task), and
sinusoidal never learns the recency rule ($0.39$). The task thus isolates
exactly the behaviour decay-based biases provide: preferring the \emph{most
recent} match requires a positional signal whose effect is relative and
monotone in elapsed time, not an absolute index.

\paragraph{Diagnostics: selective copy.}
Ordered copying from a growing blank field (Figure~\ref{fig:synth-extrap}
right) reverses the picture. Every scheme except sinusoidal and CABLE solves
the training length; CABLE fails outright ($0.17 \pm 0.04$ across seeds,
barely above the blank-guessing floor), suggesting that a purely
content-accumulated notion of distance provides no stable ordinal position
signal for ``emit the $k$-th data token'' routing. Under $4\times$
extrapolation among the schemes that do train, the bucketed T5 bias is best
(0.95), followed by NoPE (0.86); SemRF (0.73) outperforms ALiBi (0.70), while
RoPE and learned-absolute collapse below 0.10. Notably, SemRF \emph{without}
its temporal term extrapolates better still on this task (0.88), showing that
distance decay---helpful for recency---has a cost when the task requires
attending across the entire span regardless of distance. Frame-conditioning
recovers part of this cost relative to ALiBi's fixed slopes, consistent with
its design.

\paragraph{Diagnostics: associative recall.}
Associative recall exposes a dimension the other tasks do not: whether the
content-matching (induction) circuit forms \emph{at all} within a fixed budget.
The transition is sharp and seed-dependent near our budget (Section
\ref{sec:setup}). The two content-conditioned biases are the only reliable
learners: SemRF and CABLE both form the circuit in every seed
($1.000 \pm 0.000$ on the training distribution); ALiBi succeeds in two of
three seeds ($0.75 \pm 0.35$), and \emph{no other scheme ever escapes the
value-guessing plateau} ($\approx 0.26$) within 15k steps. The ablation is
decisive about the mechanism: removing SemRF's temporal term drops training
accuracy to $0.33$---content-conditioned decay is what enables reliable
circuit formation, likely by concentrating early attention locally and
breaking the symmetry among candidate key--value bindings. Under gap
extrapolation, CABLE is strongest: its accumulated ``distance'' assigns
near-zero increments to filler tokens, so retrieval is unaffected by gap
length (accuracy $1.000$ at every gap)---the clearest empirical advantage of
its cumulative-sum design, and one SemRF does not match. ALiBi's fixed slopes
are likewise scale-free and hold near-perfect accuracy when its circuit
forms, whereas SemRF's \emph{learned} temporal parameters adapt to the
training gap scale and over-decay at $20$--$40\times$ longer retrieval
distances ($0.87$ at gap 64, $0.39$ beyond); the hard-assignment variant
retains more ($0.80$ at gap 160 in its best seed). We report this
trade-off---SemRF's frame-level decay buys formation reliability,
language-modeling accuracy, and ordinal-position competence (selective copy,
where CABLE fails), at the price of gap-scale adaptation on pure
content-keyed retrieval, where CABLE's token-level accumulation excels---as a
limitation and a target for future work (e.g., parameterizing frame decay in
accumulated content rather than raw distance). In a harder training regime
(gap jitter up to 32), no baseline then evaluated solved the training
distribution in any seed, while SemRF succeeded in one of three
(Appendix~\ref{app:hard}).}

% --------------------------------------------------------------- ablations --
\newcommand{\resAblationProse}{%
Figure~\ref{fig:ablation} ablates each SemRF term, the soft assignment, and the
anchor count $K$. The clearest causal result concerns the temporal term: on
temporal recency, removing it leaves training-length accuracy intact
($0.997$) but destroys extrapolation ($0.86$ at $2\times$, $0.55$ at
$4\times$, versus $1.000$ at all lengths for the full model)---the
frame-conditioned decay is directly responsible for length-robust
time-sensitivity, not merely correlated with it. Conversely, on selective copy
the no-time variant extrapolates best ($0.88$ vs.\ $0.73$ at $4\times$),
quantifying the trade-off discussed above. Dropping the anchor-affinity or
residual terms changes selective-copy extrapolation by at most $0.03$,
indicating the content terms matter more for language modeling (where SemRF
beats ALiBi at the training length) than for these purely mechanistic tasks.
Hard (straight-through) assignment matches soft assignment on temporal recency
($1.000$) and costs a little on selective copy ($0.68$ vs.\ $0.73$); accuracy
is stable across $K \in \{8, 16, 64\}$, with mild gains at larger $K$ on
selective copy.}

% ---------------------------------------------------------------- analysis --
\newcommand{\resAnalysisProse}{%
We inspect the SemRF module of an enwik8 model (seed 0) after training.
\textbf{Anchors discover character classes.} Of $K{=}32$ anchors, 12 are used
by the 98 byte values that occur in the corpus, and the induced hard clusters
align with interpretable classes (Figure~\ref{fig:anchors}): the bulk of
lowercase letters share one frame; digits split across dedicated frames;
uppercase letters, whitespace/control characters, and punctuation each form
their own groups; and the wiki-markup delimiter \texttt{|} occupies a
singleton frame. None of this structure is supervised---it emerges from the
language-modeling objective alone.
\textbf{Frames acquire distinct time constants.} The learned per-frame decay
slopes (Figure~\ref{fig:anchors}, right) vary substantially \emph{within} each
attention head: the within-head spread between the slowest and fastest frame
reaches 88\% of the head's mean slope. The ordering is semantically coherent:
the slowest-decaying frames---those whose queries retain the longest
context---contain the structural markup tokens (\texttt{|}, punctuation,
whitespace/newline), while the fastest-decaying frames contain lowercase
letters, whose next-token prediction is dominated by word-local context. The
model thus learns, without supervision, that \emph{structure is long-range and
orthography is local}---precisely the decoupling of ``what'' from ``when'' that
motivates SemRF.}

% -------------------------------------------------------------- conclusion --
\newcommand{\resConclusionClaim}{%
delivers the best character-level perplexity of eight positional schemes at
the training length, extrapolates to $8\times$ longer contexts with
monotonically improving loss where rotary and absolute schemes degrade or
diverge, and is the only scheme---including the content-conditioned CABLE
baseline---to solve all three diagnostics at the training length, with
ablations attributing its recency and recall behaviour to the
frame-conditioned temporal decay.}

% ---------------------------------------------------------------- appendix --
\newcommand{\resAppendixProse}{%
\textbf{Synthetic tasks.} Associative recall: 16 keys / 16 values, 8 pairs,
8 queries, filler gap sampled uniformly in $[0,8]$ at training time; models
$d{=}256$, 4 layers, 8 heads, $d_{\mathrm{ff}}{=}1024$; AdamW, lr $10^{-3}$
(cosine, warmup $=$ steps$/20$), batch 64, 15{,}000 steps. Temporal recency: 4
variables, 16 values, stream length 96, query probability 0.5, 10{,}000 steps.
Selective copy: 16 data tokens in a 128-token field, 5{,}000 steps. Baselines
use three seeds and SemRF ablations two; we report mean $\pm$ population std
and, for headline comparisons, paired $t$-tests across seeds (with $n{\le}3$
these are indicative rather than conclusive; the qualitative separations in
Figure~\ref{fig:synth-extrap} do not depend on them). We verified in
preliminary probes that the content-matching circuit does not form reliably in
smaller or shorter-trained models (a sharp transition between 6k and 15k steps
at $d{=}256$ for associative recall), consistent with capacity effects
reported for MQAR-style tasks; near this transition, circuit formation is
seed- and hyperparameter-sensitive for \emph{all} methods, which is itself one
of our observations.
\textbf{enwik8.} Byte-level (vocab 256), 90M/5M/5M split; $d{=}384$, 6 layers,
6 heads, $d_{\mathrm{ff}}{=}1536$, dropout 0.1, weight tying; AdamW, lr
$6{\times}10^{-4}$ (cosine, 400-step warmup), batch 16, context 512, 10{,}000
steps ($\approx$82M tokens); bf16 autocast; evaluation over non-overlapping
windows on 500k test bytes per context length. SemRF: $K{=}32$ anchors,
residual dim 32, $\tau{=}1$, shared content terms, per-head frame slopes.
\textbf{CABLE.} We follow the authors' reference implementation
(\texttt{github.com/axiomlab/cable}, \texttt{pos\_methods/cable.py}, class
\texttt{CABLE}, which their GPT experiments use): each layer owns a
$\mathrm{Linear}(d \to H)$; per-token increments $-\mathrm{softplus}(\cdot)$
are cumulatively summed into $S$ and the additive bias is $S_i - S_j$,
recomputed at every layer from that layer's LayerNormed attention input; the
projection keeps PyTorch's default initialization, as in the reference. The
released reference omits the separate amplitude weighting $g_\theta$ of the
paper's Eq.~7, and we match the released code.
Training cost: SemRF $\approx$1.55$\times$ the per-step time of RoPE at the
training context in our implementation (additive-bias schemes use PyTorch's
math attention path rather than the fused kernel).}
