\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,booktabs,graphicx,microtype,xcolor}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{natbib}
\input{numbers.tex}

\title{Capabilities Return: What Unlearning Leaves Behind,\\ and What Warning Remains}
\author{Gunner Levi Howe\\ \small Independent Researcher\\ \small gunnerlevihowe@gmail.com}
\date{}

\begin{document}
\maketitle

\begin{abstract}
\noindent\textbf{Scope caveats first:} every experiment here is at toy scale (two- to
six-layer transformers on synthetic languages), the re-emergence arms use $n{=}3$ seeds
extended to a pre-registered $n{=}10$ confirmation, one capability family (induction /
in-context copying) carries the timing results, and one unlearning family
(data-space active unlearning) is tested. Within that scope, every claim below was
pre-registered with kill criteria before its data existed, scored by sealed one-shot
scorers against committed artifacts, and survived (or is reported as killed).

In a companion paper we showed that a capability's \emph{first} emergence can be
forecast: a mechanistic precursor (the previous-token head) leads the induction
capability by a config-invariant fraction of training, with certified false-alarm rates.
Here we ask what happens \emph{after} a capability is removed. Three findings. (1)
\emph{The precursor is a prerequisite, not the clock.} Supplying it at initialization
causally accelerates emergence \PsevenSpeedup$\times$ (complete separation,
$p={}$\PsevenExactP; placebo and near-miss priors do not accelerate), a
\emph{sub-threshold} seed captures most of the effect, and the dose--response is
U-shaped: over-supplying a converged pattern is worse than seeding a weak one. (2)
\emph{Fingerprint monitoring of a guarded model gives certified detection with a
characterized blind spot.} A library of capability fingerprints captured from an
unguarded specimen detects faithful and relocated returns with zero false alarms on
capability-blocked streams, but a return re-formed in an orthogonal representation
defeats the frozen fingerprint entirely (behavior \PilotIdioFinalAcc{} vs.\ fingerprint
readout \PilotIdioFinalStruct). (3) \emph{Unlearning removes the behavior, not the
scaffold --- so capabilities return.} After active unlearning drives copying to chance
(and in-distribution copying \emph{negative}), the precursor survives at
\RtwoScaffoldLo--\RtwoScaffoldHi; re-teaching restores the capability in
\RtwoFaithMed{} steps versus \PsevenTzero{} for first emergence
(\RtwoRelearnX$\times$ faster). Denying the surviving scaffold doubles time-to-return
(ratio \RtwoBRatio; pre-registered confirmation at pooled $n{=}10$
\ifnum\RtwoCready=1 : ratio \RtwoCRatio{}, 95\% CI [\RtwoCCIlo, \RtwoCCIhi],
\RtwoCVerdict\fi) --- there is no free disguise --- yet the denied return remains
\RtwoBDeniedVsFirstX$\times$ faster than first learning, because everything else
persists. The safety reading is uncomfortable and, we believe, decision-relevant:
behaviorally unlearned is not removed; re-exposure recovers capability on timescales
far shorter than first learning with or without the original circuit; and monitors of
guarded models get certified \emph{detection} but only fractional, not absolute,
\emph{warning}.
\end{abstract}

\section{Introduction}
A frontier lab that removes a dangerous capability from a model and then continues
training it --- fine-tuning, continual learning, tool-use adaptation --- inherits a
monitoring problem: \emph{will the removed capability come back, and can we see it
coming?} This paper studies that problem end-to-end at toy scale, where every quantity
is measurable and every intervention is clean, as the third paper in a program on
emergence timing \citep{companion2026a,companion2026d}.

The companion result this builds on: for induction --- the canonical in-context copying
capability \citep{olsson2022context} --- a mechanistic precursor (the previous-token
attention head) forms well before the behavior, leads it by a roughly config-invariant
fraction of training, and supports calibrated, false-alarm-certified forecasts of
\emph{first} emergence. The natural extension to the guarded-model problem seems
obvious: capture the capability's internal fingerprint while it exists, remove the
capability, and watch for the fingerprint's return.

We report what actually survives when that idea is subjected to the certification
discipline of the program (pre-registration with kill criteria, manufactured
capability-blocked negatives, sealed one-shot scoring). The result is a reshaped
picture:

\textbf{Part I (causality).} The precursor is causally load-bearing but is not the
clock. Supplying it at initialization accelerates emergence \PsevenSpeedup$\times$
with complete seed separation; an equally aggressive task-useless prior and a
near-miss prior (offset by one position) do nothing. A prior \emph{weaker than our own
detection threshold} captures most of the effect, and the dose--response is U-shaped
with an interior optimum near the natural formation strength: the network amplifies a
weak seed and \emph{fights} an over-strong one.

\textbf{Part II (monitoring).} A fingerprint library captured from an unguarded
specimen yields a monitor with certified false-alarm behavior --- silent on omission
and on an adversarial matched-data stream in which the capability's training signal is
present but unlearnable --- and it catches faithful returns and returns that relocate
to a different attention head. Its blind spot is representational: a return forced to
re-form orthogonally to the captured subspace restores full behavior while the frozen
fingerprint reads noise.

\textbf{Part III (unlearning and return).} Active unlearning that drives the behavior
to chance --- and beyond, to \emph{anti}-copying in distribution --- leaves the
precursor scaffold intact at \RtwoScaffoldLo--\RtwoScaffoldHi{} attention mass.
Re-emergence is therefore not re-learning: it is re-composition on surviving
infrastructure, \RtwoRelearnX$\times$ faster than first emergence. Denying the
scaffold (which requires double the bias strength that suffices during formation ---
converged structure resists development-strength interventions) restores a relative
cost: the denied return takes \RtwoBRatio$\times$ as long. But it is a relative law
only; in absolute terms the denied return is still \RtwoBDeniedVsFirstX$\times$
faster than first learning.

We also report what the discipline caught along the way: a guard that passed a naive
reading but had only unlearned the capability \emph{in distribution}; a ``disguise''
arm whose burn was silently defeated by a converged head; and an early loss-dynamics
detector (in a companion rung) that passed every standard evaluation while reading the
data mix rather than the capability. We believe the certification regime --- not any
single positive result --- is the transferable contribution.

\section{Setup and certification discipline}
\label{sec:setup}
\textbf{Substrate.} Decoder-only transformers (2 layers, $d{=}256$, 4 heads, context
256 for the timing substrate; up to 6 layers for the multi-capability monitor) trained
on synthetic Markov languages in which a fraction of sequences repeat their prefix at a
variable offset, making match-based induction the only general solution. The
behavioral event is an absolute crossing: copy advantage $\ge 2.0$ nats (the
difference in cross-entropy between first and second halves of a repeated random
probe), sustained for two consecutive evaluations. First emergence on this substrate:
median \PsevenTzero{} steps ($n{=}\PsevenCtrlN$).

\textbf{Instruments.} The precursor read is the maximum over heads of mean attention
to the immediately preceding position (previous-token score); the capability's own
circuit read is the prefix-matching score; the composed anchor --- the only alarm rule
in the program that has never false-alarmed anywhere it was scored --- is the
conjunction of the precursor and an in-distribution behavioral ramp in the same
evaluation.

\textbf{Discipline.} Every rung is pre-registered (predictions and kill criteria
committed before the data exists), scored by a sealed one-shot scorer that refuses to
run on partial fleets or twice, and adjudicated against manufactured
capability-blocked negatives: data ablation (no repeats), architectural blocking (one
layer), and adversarial matched-data streams (the capability's marker present at
identical statistics but with scrambled targets, so a detector that fires is reading
the data condition, not the capability). Manipulation checks are scored before
outcomes and can void an arm; two arms in this paper were voided and re-run this way.

\section{Part I: the precursor is a prerequisite, not the clock}
\label{sec:p7}
\textbf{Causal seeding.} We add a fixed additive attention-score bias
$B[i,i{-}1]{=}\beta$ to a single layer-0 head at initialization --- an architectural
supply of the previous-token pattern --- and measure emergence timing against
\PsevenCtrlN{} controls. At $\beta{=}8$: median \PsevenHardMed{} versus
\PsevenTzero{} (\PsevenAccelPct\% acceleration, \PsevenSpeedup$\times$), with
\emph{complete separation} (every scaffolded run beat every control; exact one-sided
permutation $p={}$\PsevenExactP). A task-useless placebo of identical magnitude
(bias to position 0) lands at \PsevenSinkMed{} and a near-miss prior ($i{-}2$) at
\PsevenNearMed{} --- neither accelerates; both are, if anything, slightly slow,
consistent with occupying the head that would otherwise become the precursor. A
no-repetition corner (scaffold supplied, training signal ablated) produces zero
events: architecture cannot manufacture the capability without data.

\textbf{The pre-registered clock hypothesis failed.} If the precursor were the sole
rate-limiter, supplying it should compress emergence to the post-anchor residual
($\sim$1{,}000 steps). Observed: \PsevenHardMed{} --- the bet missed by
$3.7\times$. The diagnosis (post-hoc, then confirmed prospectively in the dose rung's
floor bet, which \emph{fired its own kill}): emergence is gated by more than one
prerequisite, and deleting the slowest merely exposes the next. We report this as the
central interpretive lesson of Part I: \emph{interventions on named precursors
re-time emergence but do not schedule it}.

\textbf{Dose--response is U-shaped; a sub-threshold seed suffices.} Sweeping
$\beta \in \{1,2,3,4,6,8\}$ (5 seeds each): medians
\PsevenDoseAMed, \PsevenDoseBMed, \PsevenDoseCMed, \PsevenDoseDMed,
\PsevenDoseEMed, \PsevenHardMed. Two facts matter. First, $\beta{=}1$ imposes an
initial previous-token weight of \PsevenDoseAPv{} --- \emph{below} the 0.10 threshold
our own campaign uses to declare the precursor present --- and still accelerates
emergence to \PsevenDoseAMed: the network amplifies the sub-threshold hint past
threshold within 100 steps. ``Supplying the precursor'' therefore does not require
supplying a converged head; a seed the network grows itself is enough, and
``precursor present'' is better understood as a rate than a state. Second, the curve
has an interior optimum near $\beta{=}\PsevenDoseOptBeta$ and \emph{rises} toward
$\beta{=}8$: the ascending arm is seed-paired with zero reversals (15/15). Per-head
instrumentation shows a sharp adopt-versus-fight crossover: at $\beta\le3$ the
network amplifies the supplied head above its imposed value in every run; at
$\beta\ge4$ it suppresses it in every run. Too-strong priors are actively resisted.

\textbf{Placement.} The identical bias on layer 1 --- where a previous-token head
cannot feed the induction circuit --- lands at \PsevenHardLOneMed{} (no
acceleration), and the biased head never becomes an induction head. We register the
honest limit: this is consistent with the composition story but under-determined
(the arm is statistically indistinguishable from the placebos, so ``layer-1 is a bad
site for any bias'' is not excluded; the discriminating layer-1 placebo is future
work).

\section{Part II: fingerprint monitoring --- certified detection, characterized blind spot}
\label{sec:p8pilot}
We instantiate the guarded-model monitor end-to-end on a two-mechanism library
(induction, with an attention-alignment fingerprint; bracket-depth counting, with a
rank-4 interior-residual subspace fingerprint captured by one-shot ridge fit from the
frozen specimen at decode $R^2{=}1.0$). Protocol: train the unguarded specimen;
capture the library from frozen weights; guard each skill by active unlearning
(answers scrambled at matched marker statistics) until behavior and structure sit at
chance; then continually train under watch streams while reading each fingerprint's
proximity inline on fixed probe banks --- never on the training stream.

\textbf{Certified detection.} The composed alarm (structural proximity above an
absolute band set from capability-blocked baselines, AND a behavioral ramp) is silent
on every negative stream --- omission and marker-scramble, both skills --- and fires
on faithful returns. \textbf{Relocation is caught.} Burning the fingerprint's home
head forces the returning skill onto a different head; the frozen-home read misses
(\PilotRelocHome) while the max-over-heads read catches it (\PilotRelocMax).
\textbf{Orthogonal re-formation is not.} Re-teaching the counting skill while
continuously erasing the captured subspace restores full behavior
(\PilotIdioFinalAcc) while the frozen fingerprint reads \PilotIdioFinalStruct{} and
the alarm never fires --- with the negatives silent, this is a verified blind spot,
not a blind monitor. A capability can return wearing a representational disguise the
specimen fingerprint cannot see; broad reads defeat cheap disguises, not this one.

\textbf{No early warning for re-emergence at this depth.} In the pilot's substrate
the guarded skills returned within a few hundred steps and alarms were coincident
(lead \PilotFaithLeadMzero) or lagging (\PilotFaithLeadMsix{} for the subspace read)
--- which motivated Part III's move to the slow substrate where first-emergence
timing is understood.

\section{Part III: unlearning removes the behavior, not the scaffold}
\label{sec:p8r2}
\textbf{Unlearning is distribution-local (a caught manipulation, kept as a
finding).} Our first guard --- training on sequences whose repeats \emph{lie}
(copied spans permuted) --- drove in-distribution copying strongly negative while
out-of-distribution copy advantage plateaued at $\sim$1.1--1.4 nats: the model
learned ``repeats lie \emph{here}'' while generic match-and-copy survived. The
frozen validity bar rejected the guard; guard v2 adds lying repeats \emph{in the
surviving circuit's own distribution} and passes on all seeds
(copy advantage $\le \RtwoGuardCopyHi$ nats, in-distribution negative).

\textbf{The scaffold survives complete behavioral removal.} After a valid guard, the
previous-token scaffold reads \RtwoScaffoldLo--\RtwoScaffoldHi{} --- essentially the
specimen value --- on every seed. Unlearning deleted the capability's output
behavior and left its upstream prerequisite untouched.

\textbf{Hence re-emergence is re-composition, not re-learning.} Re-teaching restores
the capability in a median of \RtwoFaithMed{} steps versus \PsevenTzero{} for first
emergence: \RtwoRelearnX$\times$ faster. The composed anchor leads by
\RtwoFaithLead{} steps --- consistent across every run, but three evaluation ticks
in absolute terms. The firewall holds (zero alarms on guard-hold streams).

\textbf{Converged structure resists development-strength interventions.} Our first
``denied'' arm burned the surviving scaffold head with the same $\beta{=}8$ bias
that reliably burns \emph{forming} heads; the converged head's learned margin
defeated it (residual previous-token mass \RtwoBurnFailLo--\RtwoBurnFailHi{} at
step 0), and the arm was voided by its manipulation check. $\beta{=}16$ silences it
(step-0 mass $\le\RtwoBManipHi$). The asymmetry --- $\beta{=}8$ suffices during
formation, $\beta{=}16$ is needed at convergence --- is the deployment-relevant
mirror of Part I's adopt/fight crossover.

\textbf{The compensating law: no free disguise, but a cheap one.} With the scaffold
genuinely denied, the return must regrow a precursor elsewhere, and pays for it:
median \RtwoBDeniedMed{} steps versus \RtwoFaithMed{} faithful --- ratio
\RtwoBRatio{} at $n{=}3$%
\ifnum\RtwoCready=1
, confirmed at pooled $n{=}\RtwoCN$: pooled ratio \RtwoCRatio{} with bootstrap 95\%
CI [\RtwoCCIlo, \RtwoCCIhi] against the pre-registered decision line of 1.25
(\RtwoCVerdict; pooled faithful \RtwoCFaithMed{} vs.\ denied \RtwoCDeniedMed{}
steps; pooled hold false alarms \RtwoCHoldFA; pooled scaffold survival
$\ge\RtwoCScaffoldLo$)%
\fi
. A disguised return costs real relative time --- there is no free disguise --- but
the absolute refuge is thin: even scaffold-denied, the return is
\RtwoBDeniedVsFirstX$\times$ faster than first learning, because the language, the
downstream circuit, and everything else the capability composes with persist through
unlearning.

\section{Related work}
\label{sec:related}
Activation-clamping during training established qualitatively that supplying or
removing induction subcircuits moves the phase change \citep{singh2024needs}; their
intervention is activation-level surgery across seeds, without placebo or near-miss
controls, dose response, or timing-law tests. Data-side steering of induction-circuit
formation timing \citep{patterning2026} is the data-dual of our architectural lever.
Architectural priors that hard-wire local-shift structure --- relative-position
biases \citep{press2022alibi}, short-convolution and token-shift primitives
\citep{fu2023h3} --- are the deployment ancestors of our $\beta$-bias; none measures
capability arrival time. On the monitoring side, our fingerprints are standard
instruments (attention pattern scores \citep{olsson2022context}; probe subspaces);
the contribution is the certification regime around them. The unlearning findings
connect to the growing literature on shallow unlearning and rapid relearning
\citep{unlearningrobustness}; our contribution is mechanistic: \emph{what} survives
(the precursor scaffold, at full strength), \emph{why} return is fast (re-composition
on surviving infrastructure, quantified against first-emergence timing on the same
substrate), and what denial costs (a relative factor of $\sim$\RtwoBRatio, not an
absolute refuge).

\section{Limitations}
\label{sec:limits}
Toy scale throughout; one capability family carries the timing claims; one unlearning
family (data-space active unlearning) is tested --- weight-space methods may leave
different residues; re-emergence arms are $n{=}3$ extended to a pre-registered pooled
$n{=}10$ confirmation with the decision line frozen at the original kill threshold;
the compensating-law ratio sits at its pre-registered bar and is reported as
boundary-grade evidence; the 25-step evaluation grid limits timing resolution; the
layer-1 placement reading is under-determined; anchor lead fractions are not
conserved across arms and we make no law claim for re-emergence warning time.

\section{The kill ledger}
\label{sec:kills}
Claims that died, pre-registered bets that failed, and manipulations the discipline
caught --- reported with the same prominence as the positives: the clock hypothesis
(missed by $3.7\times$); the two-gate fixed-assembly model (killed by its own
prospective floor bet in the dose rung); ``placebos are inert'' (they delay);
the seed-stickiness arm (unreachable bar --- arithmetic, not evidence); the
loss-dynamics detector that passed every standard evaluation while reading the data
mix (killed by a matched-data capability-blocked cell); guard v1 (distribution-local
unlearning); the $\beta{=}8$ burn (defeated by a converged head); and the fingerprint
early-warning claim for re-emergence (no absolute warning exists at this scale).

\bibliographystyle{plainnat}
\bibliography{references}
\end{document}
